---
title: Root Shoot assessment in nursery tree stock

author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
csl: tree-physiology.csl
bibliography: references.bib
---

```{r, results='asis', echo=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
```


#Overview
One of the most critical statements of the "Australian standard for Tree stock and landscape use" is that 


Evaluating nursery seedling quality is necesasry to understanding seedling development and success after outplanting, however the quality of tree stock often is assessed inconsistently [@haase2008understanding].

### survival and establishment (ecology and economy)

1. lanscaping
2. orhcards
3. plantation
4. urban

What constitutes success of outplantings.  After what duration post outplanting is surivial/establishment considered complete.


# The major assessment players 
*1. Tree stock balance: relationship between 'Size Index' and Rootball Volume  
*2. Size Index (as Height x Calliper) 
*3. Height : Calliper
*4. Rootball 

Height and diameter are easily the two most common parameters examined in tree stock,
Usually set with with minimum and maximum targets established in grower contracts [@haase2008understanding].

 but what about roots
 

 It is important that the shoot not be too tall relative to the stem diameter and that the shoot mass not be too large relative to the roots [@haase2008understanding].

## Review of Root : Shoot balance 
*1. What root:shoot means to generatl tree ecology
*2. What does root:shoot mean for nursery tree stock
*3. root:shoot in bareroot vs container
*4. root:shoot across container types and/or sizes


### Aboveground (Height, Diameter/Calliper)

### Belowground (Rootball diameter and volume)

How do these root parameters affect/determine outplanting success.

## The bench warmers

### Aboveground

### Belowground

*rootball occupancy*

*root direction*
Root spiralling was found occur in all *Pinus pinea* seedlings grown in containers, however, spiralling had no affect on success following outplanting [@dominguez2006container]

*root division*
*height of root crown*  
*growin density

# Significant to the nursery industry

Allometric approaches to model nursery stock tree growth.

Larger container volumes lead to greatest field performance via incrased height, diameter and nutrient content in *Pinus pinea* seedlings, with an optimal relationship of container depth:diameter or 4 [@dominguez2006container].

#australias climate zones
 Bureau of Meteorology Climate Zones.  These are based on a modified Köppen classification system.

\
```{r ausclimate, echo=FALSE, error=TRUE,fig.width=8, fig.height=6,dev="png", dpi=600, dev.args=list(type="cairo")}
library(png)
climate <- readPNG("images/climatezone_aus1.png")
par(mar=c(0,0,0,0))
plot(c(0,1),c(0,1), type="n", axes=F, xlab="", ylab="")
rasterImage(climate, xleft = 0, ybottom = 0, xright = 1, ytop =1)

```
\

\
```{r ausclimate2, echo=FALSE, error=TRUE,fig.width=8, fig.height=6,dev="png", dpi=600, dev.args=list(type="cairo")}
library(png)
climate2 <- readPNG("images/climatezone_aus2.png")
par(mar=c(0,0,0,0))
plot(c(0,1),c(0,1), type="n", axes=F, xlab="", ylab="")
rasterImage(climate2, xleft = 0, ybottom = 0, xright = 1, ytop =1)

```
\

This raises the queston of how large of an impact does climate variability on nursery stock tree growth.

# When to assess
After outplanting or across seasons to better understand overall seedling quality[see haase]?
