---
title: Tree stock balance in nursery tree stock

author: ""
date: ""
output:
  word_document:
    reference_docx: manuscriptstyle.docx
csl: tree-physiology.csl
bibliography: references.bib
---

```{r, results='asis', echo=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
```


#Overview (the Standard and what it means)
Nursery provides this and this.  Worldwide for this.  This in Australia. Specifically, in Australai the new 20/20 vision sets goals to incrase greenspace, .. As a result, the need to 
1. lanscaping
2. orhcards
3. plantation
4. urban
economy, ecology, purpose here


Recently, the nursery industry in Australia has adopted a new standard for tree stock for landscape use (AS2303:2015). Although this standard is not currently mandantory, it was developed to 

One very important statement within the "Australian standard for Tree stock and landscape use" is that 
variability between tree stock should be expected, as trees are living organisms and there is nothing exact
in their nature. This variability presents a large challenge for the nursery industry as they should attempt to produce a uniform planting stock in which the variance of the morphological characteristics is reduced [@puttonen1997looking]. Irregardless, nursery seedlings should embody the structural and physiological traits that can be quantitatively linked to success in the field [@rose1990target]. Evaluating nursery seedling quality is thus necessary to understanding seedling development and success after outplanting, however the quality of tree stock often is assessed inconsistently [@haase2008understanding].


##shifts in nursery tree stock with 20/20/20 and new stadard in OZ.  green spaces and nursery trees
#from 2020 to 202020 (recent trends and directions of Australia nursery)
In 1997 the Australian Federal Government set a target to treble the nation’s plantation estate by 2020: the ‘2020 Vision’ (www.plantations2020.com.au). This initiative has seen the expansion of the plantation estate by 55 % to 1.97 million ha, hardwoods accounting for 48 % of the total (Gavran and Parsons 2009)....60 % Eucalyptus globulus Labill and
20 % Eucalyptus nitens Maiden.

The first 2020 vision saw a shift from bareroot to containerised production of tree seedlings [@close2012review]

Now ----20% more greenspace establishded a significant shift in the what Austrlias nurserys will do.  species for urban, new requirements?

This is highlight by the difficulty in establishment of street trees (Rob bodenstaff)

##what is quality (performance potential)

Seedling quality is the basis for tree planting success and high quality trees will have a higher survival rate and faster growth in the field than poor quality trees [@wightman1999good]. Seedling quality is a dynamic process that is a culmination of all the practives that have preceeded and will succeed that point fo measurement [@mexal1990target]. A primary goal of seedling quality assessments is to quantify levels of morphological and physiological attributes which accurately assess the condition and potential for growth and development of different stocktypes [@wilson2006quality]. As there is no one single test which encompasses seedling quality, assessing a seedling is analagous to a physician conducting a multitude of measurements to characterize a patients general health [@ritchie1984assessing].


The term "stocktype" is used to describe a seedlings age and method production, while also serving as a visual reference of what the seedling should look like before outplanting [@pinto2011conducting]. Importantly, planting seedlings with desirable plant attributes does not guarantee high survival, but does increases chances for survival[@grossnickle2012seedlings]. As seedlings are more acclimatized to nursery conditions than to planting site conditions, assessments of stocktype performance potential does include some systematic error [@puttonen1997looking]. Additionally, assessments during nursery production can be problematuc as seedling characteristics often change during the high grow phase [@mattsson1997predicting].

The ultimate goal of a generating a high quality tree stock is to ensure a very high percentage of outplanting establishment. Transplant success and establishment is dependent on the chain of events from propagation and production, to harvest and transport, to transplanting and aftercare [@struve2009tree]. To be extablished, a transplanted tree must generate a root system so that shoot growth is comparable to a non-transplanted tree [@watson1997principles]. Establishment likely different for different stock types, such as large vs small calliper trees. The ability to re-establish a balance between above and belowground growth will depend on the rate of root growth potential, even with shoot growth being typically suppressed intially [@struve2009tree]. Larger trees generally require a longer time to produce a root system compared to smaller trees [@Watson2005], which may affect outplanting success.


##Primary stresses

The three primary types of stress that influence seedling quality are moisture, temperature, and physical stress.[@haase2007morphological]. Nursery seedlings can be profoundly impacted by these primary stresses during all phases of preoduction, including culturing, lifting, packing, grading, handling, pruning, storage, and transport.

Depending on destination of the outplanting (ie landscaping, plantation forests or urban sites), seedlings will undergo varying degrees of environmental stresses not experienced during nursery production.  The varying degrees of harshness inherit by outplanted seedlings determine the length and severity of seedling 'transplant shock'.During an initial growth lag phase, reductions in water and nutrient uptake and the loss of root carbohydrates to regrowth roots will determine the amount of transplant shock and eventually outplanting success. Transplant shock are the negative effects on growth and survival when nursery-raised stock are planted into a new environment and is associated with acclimatisation of seedlings to the new environmental conditions [@close2005physiological]. During an initial growth lag phase, reductions in water and nutrient uptake and the loss of root carbohydrates to regrowth roots will determine the amount of transplant shock and eventually outplanting success.Reductions in transplant shock can thus be actively managed in nurseries through cultural practives that assess above and belowgroudn balance of tree planting stock.

***insert urban ecology, Australia and greenspace examples here
Minimizing transplant shock is highly relevant for tree stock in Australia as planned increases in urban greenspaces are underway as well as climate and soil contraints that typically define Australian ecosystems. 20/20/20 urban examples.  Aridity examples

To minimize transplant shock requires not only stocktypes that exhibit high quality morpholoical characteristic but a seedling that is properly balanced above and belowground.


##target seedling approach (variables that are most related to growth---which morph variables are most correlated to those)

focus on morphology (as a index of quality and a surrogate for physiology)

Measuring morphology in the nursery is standard practice because it easily tracks growth and describes seedlings at harvest, and has thus evolved into classification which correlates seedling survival and growth with specific morphological traits [@pinto2011morphology].  Ritchie 1984

Morphological attributes are considered a reliable measure of seedling quality as they retain their mark on the seedling identity for extended timeframes after seedlings are field planted and start to grow [@puttonen1997looking, @grossnickle2012seedlings].

Seedling morphological parameters, including height, stem diameter and shoot-root ratio each influence seedling tolerance to environmental stress and thus should to be considered in relation to each other [@cleary1978regenerating]

The morphological attributes of target seedlings represent the cumulative series of physiological processes responding to resources and stresses during nursery production [@mexal1990target].

An overarching aspect of the target seedling approach is that seedling quality is of the utmost importance.

No single morphological factor has been shown to provide a perfect prediction of outplanting success, but many of them are linked with seedling performance potential in some way [@mattsson1997predicting; @haase2007morphological]. Consequesntly, seedlings quality represents how height, diameter, plant nutrition, health, root size and shape act together and influence one another [@wightman1999good].

Height and diameter (Root collar diameter, calliper, stem diameter) are easily the two most common parameters examined in tree stock, and minimum and maximum targets are usually established in grower contracts [@haase2008understanding]. 


However, it is now commonly accepted that height and diameter measurements alone do not always correlate with seedling performance following outplanting. This realization led to the 'target seedling concept' by Rose et al. [-@rose1990target], which proposes that numerous physiolgoical and morphological seedling traits should be tracked and developed in order to quantitatively assess seedling field performance [@rose1995target]. The adaptation of this concept has led to the generation of a suite of parameters, depending on nursery practice and ...., that are currently utilized to evaluate seedling quality and potential for outplanting success.

These parameters are broadly defined into two groups, morphological and physiological, which are categorized in Table 1.

Nursery stock can be described by both morphological and physiological characteristics, but must be related to outplanting performance [@landis2011target].

Physiology and vigor can change significantly between harvest and outplanting while morphology tends not to change during that time, However, correlations stemming from morphology generally still serve as a proxy for physiology [@pinto2011morphology].

Due to a lack of a rapid and encompassing physiolgoical test [@pinto2011conducting], moprholigcal and physiological assessments are rarely conducted in combination [find more @hobbs1984influence].

Ratios of various morphological traits (e.g., root:shoot, height:diameter) have also been considered in morphological studies (Bayley and Kietzka 1997; Jacobs et al. 2006). Operational quantification of some moprhological variables are not practical for nurseries to implement on a large scale, but if superior predictors can be identified it may be possible to modify nursery cultural techniques to improve quality [@wilson2006quality]. (utilize table 1....modeify to global options)

Overall, the most widely used moprhological parameters include shoot height, root collar diameter, sturdiness (height to diameter ratio), and root to shoot ratio.

##issue with morphology and single parameter relationships (pitfalls)

Issues with using only morphological assessments, especially involving single parameter estimates of quality, have long been recognized has having overly large variation. For example, Wakeley [-@wakeley1954planting] noted how morpholigcal assessments of root collar diameter and height led to unreliable grades of survival and growth in longleaf and slash pine seedlings. 

Morphological parameters only measure overall seedling size, growth potential and shoot to root balance, not seedling physiological quality [@mexal1990target, @grossnickle2012seedlings].

Unfavorable grades of seedlings, without inferring different capacities for field success, represent a fundamental economic issue for the nursery industry. Despite this, these indices are still commonly used as the standards in modern .

Seedling quality and performance potential, however, can also vastly differ among seedlings taken from different nurseries, even when they are produced from the same seedlot over the same growing season [@pinto2011conducting].

South et al. (2005) state that ‘‘although the size of (tree seedling)
container stock has been increasing, research to show that seedling diameter is positively
related to field performance is lacking.


# Breaking down the major assessment players 
*1. Tree stock balance: relationship between 'Size Index' and Rootball Volume  
*2. Size Index (as Height x Calliper) (calliper or root collar diameter)
*3. Height : Calliper (taper development)
*4. Rootball (volume) (RGP)




# Significant to the nursery industry (building quantitaive links between parameters and root:shoot balance)

Allometric approaches to model nursery stock tree growth. Multiple regression models have been shown to better predict seedling quality than with single parameters [@jacobs2005relative], impling that combining above and bewloground indices of seedling quality will more reliably predict seedling field performance.

A cost effective sampling procedure is needed, limiting destructive sampling, as every seedling to be outplanted cannot be measured or assessed easily with current techniques [@puttonen1997looking].

#Australian standard:
little emphasis on physiolgical assesement in the nursery. The question remains on whether the morpholigcal 
indices defined in the standard represent proper root:shoot balance for the most common and widely distributed stock types in Australia.  

###do other countries evalue tree balance? in what way?
##US resources based on target seedling approach:

Commonly-measured morphological characteristics include shoot height, stem diameter, and root system size [@rose1990target](Rose et al. 1990). Tree nursery standards from various countries generally focus on these 3 core parameters to assess tree stock balance, albeit in different ways.

Americanhort: american standard for nursery stock (2014)

examples include Forestry Nursery Manuals produced through the USDA for bareroot seedlings or seedlings in containers with volumes spannning from 1989-2010.

##Europe
The European Nurserystock Association (ENA) has produced the "European technical & quality standards for nurserystock", most recently in 2010,  




## Review of Root : Shoot balance (need strong section here in terms of ecology and nurser specifics)
Combinations of root and shoot morphological characteristics may be better able to predict growth potentials and possibly outplanting success than simple shoot parameters.

*1. What root:shoot means to general tree ecology
The shoot-root ratio represents the balance between the transpirational area (shoot) and the water absorbing area (root) of a seedling [@thompson1985seedling]. 

*2. What does root:shoot mean for nursery tree stock
In nusery trees, it is important that the shoot not be too tall relative to the stem diameter and that the shoot mass not be too large relative to the roots [@haase2008understanding].

Shoot:Root ration has also been shown to be responseive to fertilzation regimes, with higher S:R with increased nutrient supply for several tree species [@villar2004nursery](Green et al., 1994; Holopainen et al.,Canham et al.,1996
1995; Graff et al., 1999), usually manifesting as increased shoot growth and not reduced root allocation. This response has also been
observed in other Quercus species.

Proper R:S balance is an important morphological attribute because it is a measure of
seedling water loss and water uptake capacity at the time of planting (Ritchie 1984;
Thompson 1985; Burdett 1990; Grossnickle 2000), although this does not always translate into reduced water stress post-planting [@lamhamed1997effect].

Higher root:shoot ratio may result in more favourable water relations,lower shoot maintenance requirements and thus faster growth rates [@close2010influences]. 

Above the optimum shoot:root ratio, survival decreases with increased allocation to shoots as evaporative surface exceeds water uptake capacity, while below optimal shoot:root ratios, decreases drought survival due to a lack of photosynthetic capacity to produce needed carbohydrate reserves [@cregg1994carbon].

Shoot-root ratios can be confounded in quality assessment when a low value does not reflects a thick taproot system instead of a large fibrous roots= system, which offers limited surface area necessary for water absorption [@ambebe2013evaluation].

South and Mitchell (1999) have cautioned that larger seedlings
do not survive better than smaller seedlings when top-growth is ‘‘out of balance’’. ===higher risk of transplant shock

Overall, a nurseries may choose to manage root:shoot balance differently when tree stock are destined to be outplanted in either arid or well irrigated environments.

*From an economic standpoint, nurseries must minimize the amount of seedlings that they destructively harvest when evaluating root shoot balance.  Thus, non-destructive morpholoigcal parameters are commonly used to assess tree stock balance. How effective these are is still a matter of contention....*

#A large question that remains is to the degree of correlation between tree stock balance and if rooting volume accurately represents the belowground status of a seedling
*3. root:shoot in bareroot vs container

*4. root:shoot across container types and/or sizes

Increassing container volume increased the root:shoot ratio at a given seedling height for Eucalyptus globulus across 10 nurseries, [@close2003physical].  Improved root:shoot ratios have been shown to be postively correlated with height growth in for seedlings [Larsen1998, others?]

multi parameter models (allows for the assessment of tree quality above and below fround )

### Aboveground (Height, Diameter/Calliper)

The first attempts to describe an quality nursery plant usually begin with morphological characteristics, such as shoot height and stem diameter, which are then converted into grading standards [@landis2006applying].

*height* 
Intro sentence..

Height is considered a good estimate of photosynthetic capacity and transpirational area, suggesting a positive relationship with subsequent growth [@haase2007morphological]. Within a nursery environment maximum shoot growth occurs at high soil water regimes and moderate to high fertility levels [@mexal1990target]. Shoot system size is important because on sites with available soil water and nutrients, competition for light between planted seedlings and the site vegetation complex is a main factor limiting seedling performance (Grossnickle 2000).Larger seedling height, however, may have adverse effects on field success in drier sites. This is mainly a issue of an imbalanced root:shoot ratio, which affects water uptake and the ability to develop new roots [cites from gross2012,@haase2007morphological], ]. Additionally, larger stock is adds difficulty in lifting, handling and planting properly, which can negate advantages of larger size tree stock in planting success [@cleary1978regenerating] 

*diameter* 
Stem diameter has been consistently found to relate to plant performance above and belowground parameters which affect seedlings growth. Diameter has also been shown to be postively related to total seedling mass and performance of outplanted seedlings for a variety of nursery grown tree seedlings [@aphalo2003field; @wilson2006quality; @zida2008initial; @bayala2009predicting].

Tree stock diameter (caliper) is tradionally viewed as a index for sturdiness for nursery tree stock [all cites].
Stem diameter at the time of planting can successfully predict of stem volume for several years after out-planting (Simpson 1995).
Stem diameter increases concomitantly with height, but this relationship is not absolute and is influenced growing density, fertility and pruning in nurseries [@mexal1990target]. nursery diameter nitrogen cites.
Diameter has been shown to be correlated with important morphological characteristics including height and dry total biomass, but not necessarily the R:S [@mexal1990target].

As stem diameter is easy to measure and is positive correlated with root system size [@cleary1978regenerating, @wightman1999good], it is an operationally attractive morphological parameter for nursery grading criteria [@dey1997morphological]. Postive linear relationship between diameter and root volume have also been noted for seedlings of northern red oak, white oak, and black cherry (Jacobs and Seifert 2004) and with red oak seedlings [@dey1997morphological].


### Belowground (Rootball diameter and volume)

Recently planted seedlings will initially depend on the root system created by nursery culture [@grossnickle2005importance], thus anticipating the potential for seedling root proliferation following transplanting could greatly improve field establishment [@davis2005quantifying]. Seedling establishment is also dependent on the capacity of seedlings to rapidly initiate new roots [@heiskanen1998influence, @grossnickle2005importance]. 

After planting the original root system size determines the ability of seedlings to take up water so they can initiate the establishment process (Carlson and Miller 1990). The size of the root system in loblooly pine seedlings, in terms of rooting volume, was shown to determine the potential for water uptake prior to new root growth [@carlson1986root].

In turn, this means that root quality parameters including ..... should be managed in order to promote seedling planting success.
New root growth will be paramount for seedlings access water and nutrient resouces following outplanting.

Many measurements of root system morphology are destructive and time consuming, which limits their application in nursery cultural practives [@jacobs2004re].

In red oak seedings seedlings, measured root volume was successful in predicting total height and diameter after one growing season [@jacobs2004re].

If rootball occupancy has meet the standard, then can container volume be used to predict aboveground growth? (will depend on the knowledge of age/tranplanting time at time of sale)

*rootball diameter*


*rootball volume*
*root growth potential (reviews by dunlap1980, simpson1997)--the quantified ability of a tree seedling
to initiate and elongate new roots within a prescribed period of time in a standard environment optimized to promote root growth (# of new roots or change in volume over standard time). Useful for seedling quality assessment but not necessarily at predicting outplanting success [@davis2005quantifying].

Folk&grossnickle: uses limtited conditions instead of standard environment to provide a more realistic test of growth potential

How do these root parameters affect/determine outplanting success.

Increases in root volume where related to increases in preplanting total fresh weight(although autocorrelated), root-collar diameter and height of Douglas-fir seedlings [@rose1991root].

Jacobs and others (2005) Second year height and diameter were greater for hardwood seedlings with greater initial root volume.

Root volume may not reflect root fibrosity, as seedlings with large fine root mass can displace the same volume as a seedlings with large tap roots [@haase2007morphological].

Development of a rapid and standardized assessment of fibrosity could help provide an accurate characterization of root system quality, as the current time consuming nature of this assessment provides little operational potential [@davis2005quantifying].

*rootball occupancy*


*rootball depth*


*root direction*
As new root must regenerate from the original outplanted root system, it is vital to assess root distribution patterns of nursery tree stock[@watson1982root].

A potential issue with the large increase in containerized seedlings is that seedlings are subject to root-binding and sprialling, which can negatively affect outplanting performance for years [@cleary1978regenerating].

Root spiralling was found occur in all *Pinus pinea* seedlings grown in containers, however, spiralling had no affect on success following outplanting [@dominguez2006container].

Root binding occurs when a plant has roots too large for its container resulting in a reduction in field performance or root growth potential, which is a constand concern for tree nurseries [@south2006root]


Generally, this issue can be avoided by root pruning, pot style...

*root division*

*height of root crown*  


*root fibrosity* (number of lateral roots representing active root tips)
*1order lateral roots*  not consistent across seedlings when predicting growth (davis cites), not enough cites and species

*root TNC* usefult to predict growth potential following outplanting as it serves as the reserve to initate new bewlogroung growth
*root nutrients* P in australia as a reserve, see Ferreira Reis and Kimmins 1986

As mentioned by Davis and Jacobs [-@davis2005quantifying], establishing standards for minimum TNC, nutrient content and moisture of the stock root system may be useful if specific to species and stocktype in terms or root system physiology. This can allow testing between root morphology and physiology with the goal of..... and which are most releveant in terms of outplanting success (belowground allometry between M and P as well as with aboveground parameters). Combine root:shoot with quality (new parameters)

###INdexs: H:D, size index, others? (dicksons), R:S

A high H:D ratio indicates a relatively spindly seedling while a lower ratio indicates a stouter seedling [@haase2007morphological].
H:D ratio was found to be correlated to mortality in Pinus patula seedlings, suggesting it may serve as a good indice of survival [@bayley1997stock]

#bareroot vs container stocktypes

Under drought conditions, seedlings in containers have higher field survival in sites with drought conditions [@grossnickle2005importance and references therein], which may have a significant role in many australian outplantings.

The root–soil contact is more disrupted in bareroot seedlings through the loss of fine roots at lifting (Nambiar 1980; Struve and Joly 1992)@rose2005root, while containered seedlings typically maintain intact multldimensional root system [@tinus1974characteristics].

Thus container grown seedlings have been commonly shown to have greater root growth during the first field growing season (cites in gross2012)[@wilson2007root]=red oak. 

Although root grown seems to be enhanced, this has not been shown to increase shoot growth and survivalbiliy in subsequent years following outplanting[@rose2005root]

Overall, container systems include better environmental control of the growing regime, shorter production cycles, increased stock uniformity and frequently superior field performance on poor quality sites (Brisette et al. 1991; Johnson et al. 1996)

Stock quality assessments show bareroot seedlings have larger shoot systems because they are typically grown at lower densities, and in many instances longer timeframes, than container seedlings [@grossnicklebareroot].

Although bareroot seedlings must additionally compete for resources belowground, the same optimal seedlings size conclusions still apply for both stock types [@aphalo2003field].

In terms of roots to shoot balance, bareroot seedlings have been found to be larger than containered seedlings with a greater S:R. [@rose2005root] Likely due to the removeal from beds.

Root morphology likely different (cites from davis2005). 
*1. containers
+ root direction
+ size and shape

The container design used for tree stocktypes has a major influence on root systems [@landis1990containers, @chapman2006early]

There are numerous container types and treatments applied to containers aimed at natural root pruning and mainpulating root direction and division.  For example, air or mechanical pruning containers and copper compounds applied to interior container surfaces are utilized in an attempt to decrease root deflection.  Container types designed to aid root pruning should produce seedlings with horizontally orientated structural roots and more stable root forms [@chapman2006early].
For example, red maple seedling height and diameter were found to similar across a large range of container types after 24 weeks, however, root deflection was decreased in containers which air or chemically pruned roots compared to standard plastic containers [@marshall1998effects]. Cork oak seedlings also had similar height and diameter after a 10 month nuresry period in shallow and deep containers, yet deeoer containers had more larger tap root and near double fine root biomass [@chirino2008effects]. Although roots deflected inside containers are commonly assocaited with tree instability, little is known about root form in large nursery containers [@gilman2010root].

*2. bareroot
+ pruning practice 
+ growing density
+ root wrenching


Based on the methods of removal, bareroot seedlings have shoot:root of 3:1 or less compared to container seedlings with a shoot:root of 2:1 or less [@haase2007morphological].

Although bareroot and container stocktypes have distinct characteristics influencing their field survival, new nursery practices are devloping bareroot seedlings with more balanced R:S and increasing large containers stocktypes [@grossnicklebareroot]. As a result, nursery tree stock standards should be more reliable between these two types..if they work at all (can be shown to be accurate)

#container volumes (surrogates for bewlogground)
Volume is one of the most obvious and important characteristics of a container as the larger the container the larger the seedling that can be produced, however, optimum container sizes can vary by species, growing density, environmental conditions and growing season length [@tsakaldimi2005root]. Container depth will determine root system growth and tap root length, which will aid in soil colonization of deep soil horizons [@chirino2008effects].

The size of containers plants are grown in prior to outplantind has been shown to have significant impacts of field success, especially during the early months after planting [@close2006defining]. For example, seedlings outplanted from very large containers may undergo water stress as large foliar water demands may outweight root uptake potential during early growth [@lamhamed1997effect].



Larger container volumes lead to greatest field performance via increased height, diameter and nutrient content in *Pinus pinea* seedlings, with a possible optimal relationship of container depth:diameter or 4 [@dominguez2006container].

not always the case....

Seedlings raised in larger volume containers may affect post-planting performance through reduced handling damage at planting, higher root:shoot ratio than smaller containers, and higher total biomass with a similar root:shoot ratio compared to plants in the different containers [@close2010influences].

Higher total biomass may result in larger total carbohydrate stores available for re-mobilisation for rapid growth soon after planting [@close2010influences].

Nelson (1996) suggested that improved after-planting performance of eucalypt seedlings produced in larger containers was due to differences in root architecture.

indicative of a
less developed root system, and can restrict the capacity for
nutrient and water uptake (Will and Teskey I997).

A review of the pot size effect found that increasing growing volume improved biomass production (Poorter et al. 2012).

The use of different containers types and volumes has been shown to have morpholoigcal consequences, including height, RCD and total mass, for black sprive seedligns [@heightpeterson1997growing].

If a container size is too small, then root restriction can will inhibit the ability of root systme to supply adequate water to shoots and will negatively affect seedling C gain [@will1997effect].

the shape and size of containers exert serious constraints on the growth of roots and their function, especially in hardwood species, adversely affecting seedling development. (Wilson et al. 2007) [@wilson2007root; @mariotti2015can]


As advanatgous as larger container volumes appear to be, this does not necessarily fit in with the economics of nursery production.  Producing high quality seedlings in smaller containers, grown at high densities, is more advanatagous to profit.  Thus, the container type can influence the economics of planting programs (check nesmith/pinto2011).

When stocktypes (Q. robur and J.regia seedlings) are grown in larger and deeper container types during nursery, height, RCD and root production, while R:S stayed the same [@mariotti2015can]

*does pot volume affect root:shoot?*
Although proper root to shoot balance will be essential for outplanting success the size of the container used for different stock types will more likely depend on nursery practives to maximum growth, yeild and profitability.  Size, time and space involved in transplanting tree stock to larger containers or dispatch them so buyers. 


Different container volumes did not affect R:S in silver birch [@aphalo2003field]

For Picea glauca seedlings S:R increased nearly two fold from 2.3 to 4 after 20 weeks from from 10cm3 - 524cm3 container volumes [@carlson1976effect].

 R:S did not shift in western larch seedlings across volumes from 111ml to 207ml  [@aghai2014container]

?why---refer to poorter

*does R:S differ significantly by species, stocktype or climate??? (KEY QUESTION)

##how management practices come into play (the impacts of container volume are obivously well studued and large, howver
##differences in nusery practices have the potentiall to.....)

Irrigation and fertilzation regimes, as well as prevailing temperatures and climate will affect seedling quality during nursery production [@mattsson1997predicting].

Nursery cultural and silvicultural practices have a strong influence on seedling performance
immediately after planting [@grossnickle2012seedlings]. This boils down to the goal of providing treestock that have the highest capacity for surivial once they leave the nursery. Seedling size in the nursery can be manipulated by the length of the growing season imposed by the timing of seed sowing [@close2006defining]. Additionally, the length of the growing season can vary across different climates zones, such as those present across Australia....what this means for evergreens, Improper nursery management may encourage a disproportionant amount of shoot growth, resulting in an unbalanced seedling with lower field-survival potential [@cleary1978regenerating].

*irrigation (amounts and hardening)



*fertilization (amounts and deprevation)

Proper fertilzation of nursery tree stock is essential for high seedlings quality, however, the degree of fertilzation may also impact the production of carbohydrates.  Alleviation of nitrogen stress on seedlings may result in less fixed carbon allocated to storage [@green1994effects; @holopainen1995resource], which may then impact the availability of starch pools for new growth following outplanting or the ability to synthesize herbivory defense compounds. In addition, nutrient defiencies (ie nitrogen or phosphorus) in nursery trees can cause decreases in leaf cholorphyll contant and the subsequent negative impact on leaf physiology can also lead to decreased carbohydrate production, tree slenderness, seedling height and RCD in nursery trees [@trubat2010nursery]. 

Commonly, the reduction of belowground resource limitation by fertilization and irrigation leads to increases in shoot growth compared to root growth, thus decreased R:S [@canham1996biomass; @jackson2012nursery] *more*
Consequently, the management of fertilization of treestock throughout the nursery period will have imporant consequences for seedling root:shoot balance, as well as future seedling establishment.

The sensitivity to the postive effects of fertilization and irrigation practives on growth rates of seedlings, however, will likely vary by species [@canham1996biomass].  Additionally, if seedlings are intentiionally nitrogen hardened before sell will reduce seedlings traints including height, rcd, leaf area and rgp [@trubat2008short]

*planting density

seedlings raised at high densities self-shade that potentially pre-disposes leaves to photodamage due
to the rapid change in exposure to light after transplanting [@close2012review].

*root pruning
Root pruning can vastly increase the surface area of the root system and increase the amount of roots within the root ball if properly managed [@watson1987effect, @gilman1996nursery]. Pruning the rootball allows for roots to grow radially straight from the trunk when planted into larger containers, decreading root morphological defects (kinks, j-rooting) [gilman2010root]. As such, root pruning is commonly performed when transplanting into containers and represents an active management practive to increase the amount of resource absorbing roots and thus seedling quality. Once outplanted, tree stability and establishment also improve with reduced root defects from active root pruning [@gouin1983girdling; @gilman2009pruning].

#effects of environment on nurseries and seedling performance

Differente environmental conditions across nursery locations can have important influences on functional traits of a common seedling type [@molla2006physiological].

ex.  affects on cold haridness (Pinus radiata and in Pseudotsuga menziesii (Menzies et al., 1981; Schuch et al., 1989)
ex. drought tolerance in meditaranian climates [
ex. possible differnces in length of growing season/daylength

q.ilex  root growth, frost resistance and drought tolerance attributed to climate in which seedlings were grown (nursery location)
[@molla2006physiological].

Different nursrey thermal regime can have an effect on survival of seedlings (Aleppo pine), but not all (Holm olk) [@pardos2003effect]. 

Often studies related on overwintering of deciduous tree stock or coastal versus inland nurseries in mediterannen climates.  In these circumstances, seedlings can become phenonlogically out of phase if dormancy is affected by temperature during over wintering.
Geogrpahical differences in nurseries will thus likely play a large role in growth of similar stocktypes, especially regarding temperature.
Mangement practives include N harderning to increase drought tolerance [@trubat2008short; @villar2004nursery]

##australias climate zones (arid zones, hot temps (urban))...how are indices affected
The cliamte of Australia means that many tree stock are container grown in open environments, and thus are exposed to .....


This potenial impact of climate has been relatively unexplored in Australia, in which tree nurseries propagate plants from tropical to temperate climates.
 
 austalia already represents an area of high aridity, nutrient soil deficiency, etc.  It is also designated as being more susceptible to extreme climate events, etc etc.  In this sense nursery standards developed to deal with an already inhospitable outplanting environment could possibly be used as a surruogate for understand the needs of other countries who are or will be facing hotter and drier climates.
 
 
 **Here sum up nursery culture and climate by saying need to include co-variates when evaluting seedling quality across Australia.  
 **Next say that age, time since transplant and species must also be accoutned for.
 
 Bureau of Meteorology Climate Zones.  These are based on a modified Köppen classification system.
 Insert figures: highlight zones around metro areas and what they represent.
 
 If differing climates affect either morphological or physiological parameters of seedling health then the extrapolation of results across sites will be less accurate.
 
 Literature: with the potential
 
 
 #how is rootshoot balance, seedling quality evaulated globally and in Australia?
Tree root to shoot balance is commonly believed to play a major role in water status of an outplanted seedling, and plays a critical role in surivival in droughted field conditions. In urban systems, this balance may critically define the success of planted tree stock as (hotter drier cite)


###What is the species effect?  Is root to shoot balance conserved?  fast, slow, native -non native, temperate decidous???








\
```{r ausclimate, echo=FALSE, error=TRUE,fig.width=8, fig.height=6,dev="png", dpi=600, dev.args=list(type="cairo")}
library(png)
climate <- readPNG("images/climatezone_aus1.png")
par(mar=c(0,0,0,0))
plot(c(0,1),c(0,1), type="n", axes=F, xlab="", ylab="")
rasterImage(climate, xleft = 0, ybottom = 0, xright = 1, ytop =1)

```
\

\
```{r ausclimate2, echo=FALSE, error=TRUE,fig.width=8, fig.height=6,dev="png", dpi=600, dev.args=list(type="cairo")}
library(png)
climate2 <- readPNG("images/climatezone_aus2.png")
par(mar=c(0,0,0,0))
plot(c(0,1),c(0,1), type="n", axes=F, xlab="", ylab="")
rasterImage(climate2, xleft = 0, ybottom = 0, xright = 1, ytop =1)

```
\

This raises the queston of how large of an impact does climate variability on nursery stock tree growth.

# When to assess
After outplanting or across seasons to better understand overall seedling quality[see haase]?


