---
title: "Bareroot Experiment @ Fleming's Nurseries"
author: "Court Campany & Leanne Gillies"
date: ""
output: html_document
---

```{r setup, results='asis', echo=FALSE, warning=FALSE}
library(knitr)
opts_knit$set(root.dir = '../')
```


```{r, echo=FALSE}
library(sciplot)
#load and formate bareroot data
bareroot2 <- read.csv("bareroot/bareroot_clean.csv")
    
# split pre and post planting data 
pre_br <- bareroot2[bareroot2$date == "2016-06-27",]
    
bareroot3 <- bareroot2[! bareroot2$date == "2016-06-27",]
bareroot3$unique_id <- with(bareroot3, paste(species, grade, sep="-"))

bareroot_agg <- doBy::summaryBy(height_m+calliper+sizeindex ~ date+grade+unique_id+genus_species+variety+species,
                          data=bareroot3, FUN=c(mean, se))
  bareroot_agg$species <- as.factor(bareroot_agg$species)
  bareroot_agg$date <- as.Date(bareroot_agg$date)
  
speclab2 <- c("Acer rubrum bowhall","Acer rubrum octoberglory","Fraxinus pennsylvanica cimmaron",
                "Fraxinus pennsylvanica urbanite","Platanus acerifolia","Platanus acerifolia liberty",
                "Pyrus calleryana capital","Pyrus calleryana chanticleer")  
```

## Summary
I split off the first set of measurements before planting in 45L. This makes sense for several reasons. It is good to have time0 measurements, which we do. It takes away the issue with caliper measurement locations before and after planting. We can analyze the two sets of data separately. We need to work on the methods. Specifically, a brief (1 paragraph) description of the bare root production process to grow the trees before we planted them. We will also need information on the soil used and the fertilization/water regime. We can add that to details below.


## Methods

##Experimental design
The experiment was located at Fleming's nursery in Monbulk, VIC, Australia. This open air wholesale tree nursery is located 220m above sea level with a mean annual temperature of 13.3 &deg; and mean annual precipitation of 1085 mm. One hundred sixty dormant bare root trees where planted into 45L bags in the winter of 2016 (July). Eight total deciduous species, consisting of two varietals per species, were planted. For each of the 16 species+varieties 10 trees of each of 2 grades ("Advanced" and "Regular") were used (Table 1). *also insert spacing/fert/irrig here*. Trees were staked as needed.

\
*insert explanation of bareroot production process and method for grading here*

\
Before planting, an initial height and diameter of each tree was measured. Initial tree height was measured starting at the root initiation point. Diameter was measured at 30cm above the root initiation point. Once tree were planted height and diameter measurements were measured at regular intervals once leaf out occurred. For planted trees, height was measured starting at the soil surface and diameter was measured at 30cm above the soil surface. The experiment was terminated on *insert date*, once a single species had reached it maximum acceptable size in the 45L container. This represents a common nursery practice in which a batch of tree stock must be either sold or potted up to the next container size.

#Results (ongoing)
Ill start running stats in the near future on the time0 trees. We will want to determine how species, varieties within species, and grades differ. For the industry, we can focus on "Size Index" if you want and/or on the allometric components height and diameter. We could also use stem volume (d^2h) as it makes a meaningful comparison to the container volume. 

##Pre-planting bareroot data

Here you clearly see how some species/varieties are inherently different. You can also see the clear difference in grade on size index.  Importantly, both height and caliper differ to start.

```{r start, echo=FALSE}
pre_br_agg <- doBy::summaryBy(height_m+calliper+sizeindex ~ date+species+grade, data=pre_br, FUN=c(mean, se))
silab <- expression(Size~index~range~~(calliper~x~height))
gradecols <- c("deepskyblue2", "firebrick3")
speclab <- c("Acer rubrum bowhall","Acer rubrum octoberglory","Fraxinus pennsylvanica \ncimmaron",
             "Fraxinus pennsylvanica \nurbanite","Platanus acerifolia","Platanus acerifolia liberty",
             "Pyrus calleryana \ncapital","Pyrus calleryana \nchanticleer")
pre_br_sort <- plyr::arrange(pre_br, species)
pre_br_sort$id <- with(pre_br_sort, paste(species, grade, sep="-"))
```


```{r start plot, echo=FALSE,fig.align='center',fig.width = 7, fig.height = 7}
par(mar=c(9,5,2,2))
boxplot(sizeindex ~ id, data = pre_br_sort,col=gradecols, xaxt='n',  ylab=silab,
        at=c(1,2, 4,5, 7,8, 10,11, 13,14, 16,17, 19, 20, 22,23), outline=FALSE)
axis(1, at=c(1.5, 4.5, 7.5, 10.5, 13.5, 16.5, 19.5, 22.5), labels=FALSE)
legend("topright", c("Advanced", "Regular"), pch=22, pt.bg=gradecols, inset=.01, bty='n', pt.cex=1.5)
text(c(2,5,8,11,14,17,20,23), par("usr")[3]-4,  srt=45,labels=speclab,xpd=TRUE, pos=2)

#Calliper
par(mar=c(9,5,2,2))
boxplot(calliper ~ id, data = pre_br_sort,col=gradecols, xaxt='n',  ylab="Calliper @ 30cm (mm)",
        at=c(1,2, 4,5, 7,8, 10,11, 13,14, 16,17, 19, 20, 22,23), outline=FALSE)
axis(1, at=c(1.5, 4.5, 7.5, 10.5, 13.5, 16.5, 19.5, 22.5), labels=FALSE)
legend("topright", c("Advanced", "Regular"), pch=22, pt.bg=gradecols, inset=.01, bty='n', pt.cex=1.5)
text(c(2,5,8,11,14,17,20,23), par("usr")[3]-1,  srt=45,labels=speclab,xpd=TRUE, pos=2)

#Height
par(mar=c(9,5,2,2))
boxplot(height_m ~ id, data = pre_br_sort,col=gradecols, xaxt='n',  ylab="Height (m)",
        at=c(1,2, 4,5, 7,8, 10,11, 13,14, 16,17, 19, 20, 22,23), outline=FALSE)
axis(1, at=c(1.5, 4.5, 7.5, 10.5, 13.5, 16.5, 19.5, 22.5), labels=FALSE)
legend("topright", c("Advanced", "Regular"), pch=22, pt.bg=gradecols, inset=.01, bty='n', pt.cex=1.5)
text(c(2,5,8,11,14,17,20,23), par("usr")[3]-.1,  srt=45,labels=speclab,xpd=TRUE, pos=2)


```

## Growth and morphology metrics
Below are the mean + standard error for each individual batch of trees. It seems that the rate of height growth in regular trees may be slightly higher and these trees seem to have caught up by February. This last date is intersting though.  It will be interesting what the new data show, as it apperas Advanced trees are leveling off. It appears that grade has a more lasting effect on caliper.  Although the rate of caliper expansion seems to be greater in Regular trees they still havent caught up by February (~20 vs ~25). Overall, Advanced trees still have a higher size index than Regular trees at this point (~70 vs ~55) due to the differences in calliper. It will be very interesting to see what the final numbers on when we terminate.

```{r sizeindex, echo=FALSE,fig.align='center',fig.width = 7, fig.height = 7}
mypal <- RColorBrewer::brewer.pal(8, "Dark2")
palette(mypal)

reg <- bareroot_agg[bareroot_agg$grade=="regular",]
adv <- bareroot_agg[bareroot_agg$grade=="advanced",]
xAT <- unique(reg$date)

# Size Index --------------------------------------------------------------
par(las=1,mgp=c(3.5,1,0),mfrow=c(2,1), mar=c(0,5,1,1))   
#plot regular trees
plot(sizeindex.mean~date, data=reg, ylab="", xaxt='n',xlab="", type='n', ylim=c(0,90))
axis.Date(1, at=xAT, labels=FALSE)

ids <- unique(reg$unique_id)
for(id in ids){
  points(sizeindex.mean ~ date, data=reg, lwd=2, 
         subset=unique_id==id, type='o', col=mypal[as.factor(species)])

with(reg[reg$unique_id==id,], arrows(date, sizeindex.mean, date, sizeindex.mean+sizeindex.se,
                                     angle=90,length=0.03, cex=2,col=mypal[species]))

with(reg[reg$unique_id==id,], arrows(date, sizeindex.mean, date, sizeindex.mean-sizeindex.se,
                                     angle=90,length=0.03, cex=2,col=mypal[species]))
}
legend("bottomright", expression(bold("Regular")), inset=.01, bty='n')
mtext(silab,2,line=-2.5, outer = TRUE, at=.5, las=0)
legend("topleft", legend = speclab2, lty=1,lwd=2, col=palette(),bty='n', cex=.8)
#advanced
par(mar=c(5,5,0,1))
plot(sizeindex.mean~date, data=adv, ylab="", xaxt='n',xlab="", type='n', ylim=c(0,90))
axis.Date(1, at=xAT)

ids <- unique(adv$unique_id)
for(id in ids){
  points(sizeindex.mean ~ date, data=adv, lwd=2, 
         subset=unique_id==id, type='o', col=mypal[as.factor(species)])
  
  with(adv[adv$unique_id==id,], arrows(date, sizeindex.mean, date, sizeindex.mean+sizeindex.se,
                                       angle=90,length=0.03, cex=2,col=mypal[species]))

  with(adv[adv$unique_id==id,], arrows(date, sizeindex.mean, date, sizeindex.mean-sizeindex.se,
                                       angle=90,length=0.03, cex=2,col=mypal[species]))
}
legend("bottomright", expression(bold("Advanced")), inset=.01, bty='n')

```


```{r height, echo=FALSE,fig.align='center',fig.width = 7, fig.height = 7}
par(las=1,mgp=c(3.5,1,0),mfrow=c(2,1), mar=c(0,5,1,1))   
#plot regular trees
plot(height_m.mean~date, data=reg, ylab="", xaxt='n',xlab="", type='n', ylim=c(0.5,3))
axis.Date(1, at=xAT, labels=FALSE)

ids <- unique(reg$unique_id)
for(id in ids){
  points(height_m.mean ~ date, data=reg, lwd=2, 
         subset=unique_id==id, type='o', col=mypal[as.factor(species)])
  
  with(reg[reg$unique_id==id,], arrows(date, height_m.mean, date, height_m.mean+height_m.se, 
                                       angle=90,length=0.03, cex=2,col=mypal[species]))
  
  with(reg[reg$unique_id==id,], arrows(date, height_m.mean, date, height_m.mean-height_m.se, 
                                       angle=90,length=0.03, cex=2,col=mypal[species]))
}
legend("bottomright", expression(bold("Regular")), inset=.01, bty='n')
mtext("Height (m)",2,line=-2.5, outer = TRUE, at=.5, las=0)
legend("topleft", legend = speclab2, lty=1,lwd=2, col=palette(),bty='n', cex=.8)
#advanced
par(mar=c(5,5,0,1))
plot(height_m.mean~date, data=adv, ylab="", xaxt='n',xlab="", type='n', ylim=c(0.5,3))
axis.Date(1, at=xAT)

ids <- unique(adv$unique_id)
for(id in ids){
  points(height_m.mean ~ date, data=adv, lwd=2, 
         subset=unique_id==id, type='o', col=mypal[as.factor(species)])
  
  with(adv[adv$unique_id==id,], arrows(date, height_m.mean, date, height_m.mean+height_m.se, 
                                       angle=90,length=0.03, cex=2,col=mypal[species]))
  
  with(adv[adv$unique_id==id,], arrows(date, height_m.mean, date, height_m.mean-height_m.se, 
                                       angle=90,length=0.03, cex=2,col=mypal[species]))
}
legend("bottomright", expression(bold("Advanced")), inset=.01, bty='n')

```


```{r diameter, echo=FALSE,fig.align='center',fig.width = 7, fig.height = 7}
par(las=1,mgp=c(3.5,1,0),mfrow=c(2,1), mar=c(0,5,1,1))   
#plot regular trees
plot(calliper.mean~date, data=reg, ylab="", xaxt='n',xlab="", type='n', ylim=c(10,35))
axis.Date(1, at=xAT, labels=FALSE)

ids <- unique(reg$unique_id)
for(id in ids){
  points(calliper.mean ~ date, data=reg, lwd=2, 
         subset=unique_id==id, type='o', col=mypal[as.factor(species)])
  
  with(reg[reg$unique_id==id,], arrows(date, calliper.mean, date, calliper.mean+calliper.se, 
                                       angle=90,length=0.03, cex=2,col=mypal[species]))
  
  with(reg[reg$unique_id==id,], arrows(date, calliper.mean, date, calliper.mean-calliper.se, 
                                       angle=90,length=0.03, cex=2,col=mypal[species]))
}
legend("bottomright", expression(bold("Regular")), inset=.01, bty='n')
mtext("Calliper at 30cm (mm)",2,line=-2.5, outer = TRUE, at=.5, las=0)
legend("topleft", legend = speclab2, lty=1,lwd=2, col=palette(),bty='n', cex=.8)
#advanced
par(mar=c(5,5,0,1))
plot(calliper.mean~date, data=adv, ylab="", xaxt='n',xlab="", type='n', ylim=c(10,35))
axis.Date(1, at=xAT)

ids <- unique(adv$unique_id)
for(id in ids){
  points(calliper.mean ~ date, data=adv, lwd=2, 
         subset=unique_id==id, type='o', col=mypal[as.factor(species)])
  
  with(adv[adv$unique_id==id,], arrows(date, calliper.mean, date, calliper.mean+calliper.se, 
                                       angle=90,length=0.03, cex=2,col=mypal[species]))
  
  with(adv[adv$unique_id==id,], arrows(date, calliper.mean, date, calliper.mean-calliper.se, 
                                       angle=90,length=0.03, cex=2,col=mypal[species]))
}
legend("bottomright", expression(bold("Advanced")), inset=.01, bty='n')

```

